---
title: "Day 1 Practice"
author: "Stephen Skalicky"
output: html_document
---

# Take home practice

This is an opportunity for some additional practice with data reading, summarising, manipulation, and plotting data.

This data is a series of plausibility ratings provided for different short stories in English. Here are the instructions participants received:

*Plausibility is a measure of how likely you think something can occur. For example, being full after eating a meal is more plausible than being hungry after eating a meal. However, one could still feel hungry after eating a meal. It is just less plausible when compared to being full.” Your task is to rate each story you read on a scale of plausibility. The scale ranges from 0 (not at all plausible) to 5 (highly plausible). When making your ratings, please consider all aspects of the story.*


The stories appeared in four different conditions - A, B, C, or D. Your goal is to write a notebook which can:

1. Load in the data file and save it to the variable `dat`
2. Create a data frame summarising the mean and standard deviation of the dependent variable
3. Create a boxplot of the ratings among the different conditions, interpret the boxplot
4. Fit a multilevel regression model comparing the dependent variable among the condition
5. Use `emmeans` to obtain pairwise contrasts, and explain the results

The data is located in the file `day-01-practice-data.csv`

# load in the libraries here
```{r}
library(tidyverse)
library(lme4)
library(emmeans)
```

## 1. Read in the data file and save it to the variable `dat`

```{r}
# add your code here

```


## 2. Create a df with mean and standard deviation of the ratings in the four different conditions. Name this df `descriptives` and create two new variables:

  - `mean_rating` = mean of rating
  - `sd_rating` = sd of rating
  
Creative your new df from `dat`, then pipe into a `group_by()`, and then pipe into a `summarise()` 

```{r}
# add your code here
```


## 3. Plot the raw data

Create a ggplot of `dat`. The y axis is rating, and the x axis is condition.

Plot a boxplot (`geom_boxplot()`) and also the raw points with jitter added (`geom_point(position = position_jitter())`). 

Style the plot to your liking. 

```{r}
# add your code here

```
## 4. Describe the plot

  - which condition(s) has the highest median rating?
  - which condition(s) has the lowest median rating?
  - where do you expect to see significant differences among these condition? Why?
  


## 5. Create a regression model predicting ratings

`rating` is your dependent variable
`condition` is your only predictor variable
add a random intercept of `(1|subject)`

name the model `m1`


```{r}
# add your code here
```

## 6. Decompose the model using `emmeans`

call `emmeans()` on `m1` and ask for `pairwise~condition`

What do you see? Write a paragraph explaining the results

- which contrasts are significant?
- among these contrasts, which is the strongest effect? Which is the smallest (hint: compare the size of the estimates)

```{r}
# add your code here
```


# bonus - what about the items? 

You'll notice there is also a variable for the items in the data - each item appeared in each condition. It might be worthwhile to see if some items attracted higher/lower ratings in general.

- Create a df named `item_summary` which is the mean and standard deviation of rating for each condition and each item. You need to add `story_id` to the `group_by()` function

```{r}
# add your code here

```


- Create a ggplot of `item_summary`. Set the y axis to be the mean rating, and the x axis to be condition. Add a facet_wrap by story `facet_wrap(. ~ story_id)`

Knowing what you know about the significant contrasts obtained among the conditions, which stories follow and which do not follow the significant effect obtained from the model? 


```{r}
# add your code here
```